---
title: "VEMCODataMgmt: Working with Passive Acoustic Data"
subtitle: "A user guide for cleaning and managing large databases of acoustic detections"
author: "Brian Moe"
date: "`r Sys.Date()`"
output: 
  github_document:
    toc: true
---

```{r, include=FALSE}
library(knitr)
library(VEMCODataMgmt)
library(data.table)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  eval = FALSE # This replaces your 'execute: eval: false' rule
)
```

This README guides the user through processing new passive acoustic detection data. In it, we walk through cleaning and standardizing new FACT network detections and merge them with the existing database. We then do the same with newly downloaded FWC Charlotte Harbor detections and merge them with the existing database. Once both new FACT and new FWC data are merged with the existing database (FWC, FACT, and iTag), we run a final quality control to ensure no repeat detections are present.

All correction rules are define in reference files stored within the "Reference Files/" directory. If working in the "Acoustic Database R Project", this will already be set up correctly. While we standardize the data and remove erroneous detections, additional quality control measures such as removing single detections (unless obviously erroneous) are not carried out at this stage.

Workflow:

1.  Install (if not already done) and load the package [`VEMCODataMgmt`](https://github.com/Brian-J-Moe/VEMCODataMgmt)
2.  Set directory paths (csv_dir and reference_dir)
3.  Clean and standardize new FACT data
    i.  Review validation report and update reference files
    ii. Rerun the newly cleaned and merged FACT data if reference files were updated
    iii. Post-processing quality control checks
4.  Clean and standardize newly downloaded FWC Charlotte Harbor detections
5.  Merge the new FACT data and new FWC Charlotte Harbor data with the existing database
6.  Quality control sanity checks

## Load the R package VEMCODataMgmt

```{r}
# If VEMCODataMgmt is not already installed:
install.packages("pak") # skip if `pak` is already installed
pak::pak("Brian-J-Moe/VEMCODataMgmt") 

# Load the package
library(VEMCODataMgmt)
```

## Clean and Standardize new FACT data

The following code is used to clean and standardize the new FACT data. In this step, we identify likely erroneous detections (e.g., detections in Canadian waters) and remove any duplicate detections that the FACT Network identifies as "new". Given the recent push for iTag users to transition to the FACT Network, it is likely that some portion of the "new" detections are already in the database.

### File path configuration for `VEMCODataMgmt`

The below configurations assume that you are working in the "Acoustic Database R Project" directory. If working in a different project which does not house a "Reference Files/" or "Data/" subdirectory within the main project directory, the full file path must be used to locate those directories.

```{r}
# Path to new FACT CSV exports folder (REQUIRED - changes with each download)
csv_dir <- "G:/Charlotte Harbor Common/Sawfish/Long Term_InternalTags/Data Received/FACT/April 2026 Dump"

# Path to reference files directory (contains correction look-up tables)
reference_dir <- "Reference Files"

# Path to existing FACT RData object and where to store the updated version
existing_db <- "Data/current_FACT_detections.RData"
output_db   <- "Data/updated_FACT_data.RData"

# Import tagging data
tagging_data <- data.table::fread("Data/Acoustic Tagged Fish.csv")
```

*Note that when loading the "Data/UPDATED_FACT_detections.RData" data file, the object that is loaded into the environment is named `FACT_detections`. To manually load and inspect this object run:*

```{r}
utils::data("UPDATED_FACT_detections")
FACT_detections
```

```{r}
#| eval: true
#| echo: false

library(data.table)

csv_dir <- "example_folder/FACT csv files"
reference_dir <- "example_folder/Reference Files"
existing_db <- "example_folder/Data/example_FACT_data.RData"

output_db <- "example_folder/Data/updated_FACT_data.RData"

tagging_data <- data.table::fread("example_folder/Data/Acoustic Tagged Fish.csv")

load("example_folder/Data/example_FACT_data.RData")

FACT_detections

```

```{r}
#| eval: true

class(FACT_detections)
```

*The column names seen above in the data object `FACT_detections` are the standardized column names used by `VEMCODataMgmt` and do not necessarily reflect those used by the FACT Network. This ensures smooth and efficient merging of data regardless of if the FACT Network changes their standard column names.*

### Execute FACT workflow

The function `VEMCODataMgmt::process_fact_workflow()` is a complete data processing and cleaning function. To view the help file for details on function arguments, run `?process_fact_workflow()`.

```{r}
#| eval: true
result <- VEMCODataMgmt::process_fact_workflow(
  csv_dir = csv_dir,
  reference_dir = reference_dir,
  existing_rdata = existing_db,
  output_rdata = output_db,
  datetime_col = "dateCollectedUTC",
  transmitter_col = "tagName",
  station_col = "station",
  lat_col = "decimalLatitude",
  lon_col = "decimalLongitude",
  poc_col = "contactPOC",
  pi_col = "contactPI",
  validate = TRUE,
  save_validation = TRUE,
  progress_bar = FALSE
) |> suppressWarnings()

```

[***The column names listed in the code above (also set as the default names) reflect those used by the FACT network as of the date this article was written. Be sure to check the new data files to be imported and modify those column names if needed.***]{style="font-size: 0.8em;"}

[*Note: Recent FACT data dumps contain "new" detections which have already been sent in previous dumps. Any changes to previously uploaded FACT data (e.g., new list of PIs or POCs) will cause the FACT database to treat them as `new` detections and include them in the next data dump.*]{style="font-size: 0.8em;"}

The output of `process_fact_workflow()` is a list object containing:

1.  The updated dataset of fact detections
2.  The newly added FACT detections
3.  Statistics pertaining to the merger of the new and existing datasets
4.  Results from the validation process

```{r}
#| eval: TRUE
names(result)
```

When `output_rdata = NULL`, the updated FACT dataset is saved in the root directory of the `existing_rdata` path under the data file name `UPDATED_FACT_detections.RData`.

Running `process_fact_workflow()` automatically imports all `.csv` files housed in the assigned `csv_dir` directory. Files are then combined and scrutinized for duplicate detection data.

*NOTE: If `csv_dir` is housed in the shared drive, you'll likely see a series of the following warning:*

```         
Warning in file.info(file): cannot resolve owner of file 'FACT csv
files/fwcsaw_matched_detections_2025.csv': No mapping between account names and
security IDs was done
```

*This warning is an artifact of using a shared network drive and can be disregarded.*

Within the subdirectory `Reference Files`, there are five `.csv` files with which `process_fact_workflow()` uses as reference files to clean and standardize the data. These files **must** remained named as follows to prevent the `VEMCODataMgmt` functions from breaking.

1.  "unwanted_agencies.csv" contains a list of known PIs which consistently provide false detections (e.g., Canadian institutions or groups). All data associated with these PIs are removed
2.  "agency_look-up.csv" is a list of past PIs and their associated agencies. This is use for standardizing the names of the agencies owning each receiver. Not that some of these pairings may be outdated. This is primarily used for consistency with past data to prevent the same data points being associated with multiple agencies. These pairings can be changed, however, some PIs upload "new" data which we already have, primarily from iTag data. If you change the agency label associated with a specific PI, post processing of the newly updated FACT data will be required to ensure no duplicate detections are present (though the validation step in the function should streamline this).
3.  "station_agency_reassign.csv" serves a similar purpose "agency_look-up.csv" and expressly looks for stations in which PIs have changed, the listed order of PIs has changed, etc., and restores them to the canonical "Agency" label.
4.  "station_name_corrections.csv" cross-references the coordinate pairs and reassigns canonical station names to those with different names. This is particularly important for some groups who regularly switch spelling or capitalization of their stations. If changes are made by PIs to the station names in the FACT database, FACT, for whatever reason, seems to treat those data as "new" and includes them in the next data dump. Additionally, it seems to be common for other groups to rename the station when redeploying a receiver or deploying a new receiver in place of the existing receiver.
5.  "Master_RECEIVERS.csv" cross-references station names with their coordinate pairs, and returns any discrepancy to the canonical coordinates. Variations in coordinate pairs often present themselves as differences in the number of decimal points or slight changes to the exact coordinate positions upon redeployment.

Once the new FACT data is cleaned, standardized, and combined, the function loads in the existing `FACT_detections` dataset currently housed within the project. Before merging the newly imported data with the existing data, a copy of the existing `FACT_detections` is saved within the "Data/" subdirectory as `old_FACT_detections`. This offers a fallback in the event something occurs when merging the two data sets and ensures there is always an available "non-updated" copy of the FACT data. It is also useful as a post-processing cross-reference tool. To load and view the non-updated data following the completion of `process_fact_workflow()`, run the console commands:

```{r}
utils::data("old_FACT_detections")
View(old_FACT_detections)
```

Once the new and existing data are merged, the function removes duplicate entries and conducts a validation check. In this validation check, the function identifies stations with the same coordinates but different names and/or assigned agency labels as well as stations with the same name but different coordinates and/or agency labels. A diagnostic table is automatically printed with function argument `verbose = TRUE` (which is the default value).

Following the validation check, the function then saves the newly merged `FACT_detections` in the "Data/" subfolder as "UPDATED_FACT_detections.RData". It also identifies all new (non-duplicate) data from the imported FACT `.csv` files, places them in a data object named `new_data`, and saves that object in the "Data/" subdirectory under the file "new_FACT_detections.RData". To view all newly added data run the console commands:

```{r}
utils::data("new_FACT_detections")
View(new_data)
```

Additionally, `process_fact_workflow()` updates the reference file "MASTER_RECEIVERS.csv" with a new list of receiver names and their corresponding `Latitude`, `Longitude`, and `Agency`. To view this file in R run the console command `View(data.table::fread("Reference Files/MASTER_RECEIVERS.csv"))`, or open and view in excel. This allows easy identification of any new PIs (the `Agency` listed will appear as the PI's full name, followed by an email address). New PIs for which the detections are erroneous can then be added to the "unwanted_agencies.csv" reference file, and any new PIs deemed to have legitimate detections can be added to the "agency_look-up.csv" reference file and given a standardized `Agency` label for future reference.

### Post Processing Quality Control Checks

`process_fact_workflow()` will not be able to identify new data quality issues not previously defined in the reference files. As such, the output dataset should be manually inspected for erroneous detections or other errors/mistakes. Below are several helpful manual checks.

View the ranges of the `Latitude` and `Longitude` columns. This will help identify false detections from obviously illogical locations (e.g., Canadian waters).

```{r}
#| eval: true

load(output_db)

FACT_detections[, .(
  lat_range = range(Latitude),
  lon_range = range(Longitude)
)]

```

Check for new PIs that are not already in the reference files. FACT formats the PI field as a single or list of individuals with their names and contact emails. The easiest way to search through the data is to identify data with the 'at' symbol (\@) within it's `Agency` text string.

```{r}
#| eval: true

# Check Agency column for new PIs
FACT_detections[grepl("@", Agency), Agency] |> unique()

```

Following any updates to the reference files, the new database can be re-run through `process_fact_workflow()` to correct for any changes.

```{r}
# Correct any changes made to the reference files by passing ONLY the new output data file path and the reference directory file path.
revised_FACT_detections <- VEMCODataMgmt::process_fact_workflow(
  csv_dir = NULL,
  reference_dir = reference_dir,
  existing_rdata = output_rdata
)

```

In the above code, only two fields are required: `reference_dir` and `existing_rdata`. When the `csv_dir` argument is missing, the function only evaluates the dataset located by the `existing_rdata` file path. Because the newly updated data was automatically saved to the specified `output_rdata` path, the `existing_rdata` path in this run should be the same as `output_rdata` from the previous run (or wherever the `.RData` file of interest is stored), in this example it is `"Data/UPDATED_FACT_detections.RData"`. The remaining function arguments are not called when `csv_dir` is missing and can therefore be left out. Unlike the output when merging new data as detailed above, when only cleaning an existing database, the output is a single `data.table` object as opposed to a list of objects.

### Additional details

`process_fact_workflow()` is a wrapper function which links together a series of functions housed within `VEMCODataMgmt`, each having their own help documentation which can be called with the console command `?VEMCODataMgmt::name_of_function()`:

1.  `VEMCODataMgmt::import_fact_csv()`
2.  `VEMCODataMgmt::process_fact_agencies()`
3.  `VEMCODataMgmt::apply_fact_corrections()`
4.  `VEMCODataMgmt::merge_fact_databases()`
5.  `VEMCODataMgmt::validate_fact_database()`

These functions were build using the `data.table` R package. As such, the resulting dataset outputs are formatted as `data.table` objects. To save the `process_fact_workflow()` output data object as a `.csv` file, use the function `data.table::fwrite(result$fact_detections, file = "your_file_path.csv")`.

***All code and associated help files can be found in the GitHub repository [Brian-J-Moe/VEMCODataMgmt](https://github.com/Brian-J-Moe/VEMCODataMgmt)***

## Clean and Standardize new FWC Charlotte Harbor data

The following code is used to clean and standardize the new Charlotte Harbor data. This step is far less arduous than cleaning FACT data and is largely for ensuring formatting of the data is consistent with that of the existing database. This produces three primary datasets:

1.  A "raw" database of ALL accumulated FWC Charlotte Harbor owned data (our tags, orphan tags, spatial reference and synchronizing tags, etc.).
2.  A cleaned and standardized database of all detections associated with our tagged animals (regardless of species)
3.  A database of detections associated with known tags deployed by FSU/NOAA

### Importing new data

Before importing and cleaning the new data, we'll want to organize the receiver logs into folders based on the date of download. This allows for efficient cataloging and handling of data and the ability to easily find download logs from specific dates (very useful for finding VPS files needed by Innovasea). First, define the file paths to specific directories and data locations. For this example, all folders with necessary files have been moved into the "README Tutorial" folder to more efficiently define the directory paths. If the folder of interest in housed within the project root directory, only the folder name is required. When calling folders outside the root directory, the full file path is required.

```{r}
#| eval: true
#| echo: false


vrl_dir <- "example_folder/Receiver Logs"
out_dir <- "example_folder/Logs by Date"
```

```{r}
# Directory containing new VRL/CSV files 
vrl_dir <- "Receiver Logs"

# Directory for which date specific folders should be placed
out_dir <- "Logs by Date"

# If a folder has not yet been created, the following code creates it within the rood directory of the project or R session:
#dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
```

Given that receiver logs for a given year are all housed within a single folder, we can define a date (`after_date`) for which the sorting function only evaluates logs downloaded after that date.

```{r}
#| eval: true

# Optional: Filter downloads after this date (NULL = process all files)
# Format: "YYYY-MM-DD"
after_date = after_date <- "2026-02-01"
```

Once both directories have been defined and an `after_date` set, use the functions `sort_VRL_logs()` and `load_detections_by_dates()` to sort newly downloaded logs into their appropriate folders and import the new detections.

```{r}
#| eval: true

# Sort newly downloaded receiver logs
sort_result <- VEMCODataMgmt::sort_VRL_logs(vrl_dir, out_dir, after_date = after_date)

# Import new data
new_det_full <- VEMCODataMgmt::load_detections_by_date(out_dir, after_date = after_date, verbose = FALSE)


# View new data
new_det_full

```

These data represent ALL transmitters detected including orphan tags, time synchronization tags, and spatial reference tags. Using a list of deployed tags, extract only those data.

```{r}
#| eval: true


# Standardize transmitter names
tagging_data[, Transmitter := as.character(Transmitter)] # ensures all IDs are character vectors
tagging_data[!is.na(Transmitter), Transmitter := trimws(Transmitter)] # Removes accidental whitespace (e.g., trailing blank spaces)


# Remove known deceased fish if desired
tags_to_remove <- c("A69-9001-54949", "A69-9001-60330", "A69-9001-54939")
tagging_data <- tagging_data[!Transmitter %in% tags_to_remove]


# Extract FWC transmitter data
new_det <- new_det_full[Transmitter %in% tagging_data$Transmitter]


```

### Post processing and data summarizations

The following example code demonstrates how to correct for a receiver initialized with the wrong coordinates (this uses `data.table` formatting).

```{r}
new_det[`Station Name` == "Glover Bight Creek", `:=` (
  Latitude  = 26.54010,
  Longitude = -81.9914
)]

```

It's also useful to generate summary statistics for all new data, specific arrays, or individual fish. The following examples walk through those processes when working with a `data.table`.

```{r}
#| eval: true


# Summarize all new detections for each fish detected
overall_summary <- new_det[, .(
  n_detections = .N,                          # total number of detections
  n_stations = uniqueN(`Station Name`),       # total number of stations detected on
  first_det = min(`Date and Time (UTC)`),     # first detection
  last_det = max(`Date and Time (UTC)`)       # last detection
), by = .(Transmitter)][order(Transmitter)]   # group by "Transmitter" and set logical order

overall_summary
```

```{r}
#| eval: true

# Summarize by grouped stations or arrays
glover_bight = c(paste0("GB", 1:19), "Glover Bight", "Glover Bight Creek")
gb_det <- new_det[`Station Name` %in% glover_bight]
gb_summary <- gb_det[, .(
  n_detections = .N,
  n_stations = uniqueN(`Station Name`),
  first_det = min(`Date and Time (UTC)`),
  last_det = max(`Date and Time (UTC)`)
), by = .(Transmitter)][order(Transmitter)]

gb_summary
```

```{r}
#| eval: true

# Summarize for specific fish of interest
favorite_fish <- list(
  "Pumpkin"       = "A69-9001-57441",
  "Helga"         = "A69-9001-42821",
  "Big Mamma"     = "A69-9001-42829",
  "Small Fry"     = "A69-9001-54931",
  "Lone Wanderer" = "A69-9001-12369"
)

favorite_results <- list()
  for (fish_name in names(favorite_fish)) {
    trans_id <- favorite_fish[[fish_name]]
    
    if (trans_id %in% new_det$Transmitter) {
      fish_summary <- new_det[Transmitter == trans_id, .(
        n_detections = .N,
        first_det = min(`Date and Time (UTC)`),
        last_det = max(`Date and Time (UTC)`)
      ), by = .(`Station Name`)][order(-last_det)]
      
      favorite_results[[fish_name]] <- fish_summary
    } else {
      favorite_results[[fish_name]] <- NULL
    }
  }

favorite_results
```


